---
title: "homework_3"
format: html
editor: visual
---

## Part 1: Build the Dataset

Create a data frame with 365 rows representing daily air quality readings at a monitoring station. Your data frame must include the following columns:

1.  `day` ŌĆö a sequence of integers from 1 to 365

2.  `pm25` ŌĆö daily PM2.5 concentration (┬Ąg/m┬│), sampled from a distribution of your choice using one of the data generation functions covered in class (`runif`, `rnorm`, `rpois`). Choose parameters that produce realistic values ŌĆö look up typical urban PM2.5 ranges to guide your choices.

3.  `season` ŌĆö a factor column with levels `"winter"`, `"spring"`, `"summer"`, `"fall"`, assigned based on day number (days 1ŌĆō90, 91ŌĆō180, 181ŌĆō270, 271ŌĆō365)

4.  `risk` ŌĆö an empty column initialized to `NA` ŌĆö you will fill this in Part 2

```{r}

library(dplyr)
library(tidyverse)
#creating data 
day = 1:365
day
pm25 = runif(min=0, max=50, n=365)
pm25

risk = NA


seasons <- case_when(
  day %in% 1:90 ~ "winter",
  day %in% 91:180 ~"spring",
  day %in% 181:270 ~ "summer",
  day %in% 271:365 ~ "fall")
seasons

seasons <- as.factor(seasons)

data <- data.frame(seasons = seasons,
           pm25 = pm25,
           day = day, 
           risk = risk)
data
```

## **Part 2 ŌĆō Classify Risk with a For Loop**

Using a `for` loop, fill in the `risk` column for each day based on the following PM2.5 thresholds (adapted from EPA AQI guidelines):

-   `"good"` ŌĆö PM2.5 below 12 ┬Ąg/m┬│

-   `"moderate"` ŌĆö PM2.5 between 12 and 35.4 ┬Ąg/m┬│

-   `"unhealthy"` ŌĆö PM2.5 above 35.4 ┬Ąg/m┬│

*Important* while there are many ways to do this - you must use a *for* loop

Within the same loop, also track the longest consecutive streak of `"unhealthy"` days. This will require checking the previous rowŌĆÖs risk classification at each step. Store the final streak length in a variable called `max_unhealthy_strea`

```{r}

min(pm25)
max(pm25)

# setting a named variable for the threshold makes the code easier to read
unhealthy = 35.4
moderate = 12
good = 0

for (i in 1:length(data$pm25)) {
  data$risk[i] <- 
    ifelse(data$pm25[i] > unhealthy, "unhealthy", 
           ifelse(data$pm25[i] >= moderate, "moderate", "good"))
  if (data$risk[i] == "unhealthy") {
    if (i == 1 || data$risk[i - 1] == "unhealthy") {
      current_streak <- current_streak + 1
    } else {
      current_streak <- 1 
      }
  } else {
    current_streak <- 0 
  }
  if (current_streak > max_unhealthy_streak) 
        max_unhealthy_streak <- current_streak 
  }

max_unhealthy_streak

data



```

## **Part 3 ŌĆō Seasonal Summary with a For Loop**

Write a `for` loop that computes number of days in each season that fall into unhealthy risk category

```{r}
# setting a named variable for the threshold makes the code easier to read
day_unhealthy <- c(winter = 0, spring = 0, summer = 0, fall = 0)


for (i in 1:length(data$day)) {
  if(data$risk[i] == "unhealthy") {
    seasons_i <- data$seasons[i] 
    day_unhealthy[seasons_i] <- day_unhealthy[seasons_i] + 1
  }
}

day_unhealthy 

```

## **Part 4 ŌĆō Visualize and Report**

Produce at least two plots:

1.  A time series or histogram showing PM2.5 concentration across the year, colored or faceted by season

2.  A bar chart showing the count of days in each risk category,

Each plot must have a title generated using `sprintf` that includes the the max_unhealthy_streak value you calculated in Part 2.

```{r}
library(ggplot2)

plot_title_A <- "Seasons"


pm25_annual_season <- ggplot(data, aes(x=day, y=pm25, color=seasons)) + 
  geom_line() + 
  labs(colour = "Seasons", 
       title = sprintf("Sampled PM 2.5 Concentrations: %d Day Unhealthy Air Quality Index Streak", max_unhealthy_streak), 
       x = "Days", 
       y = "PM 2.5 (┬Ąg/m┬│)") +
  scale_color_discrete(labels = c("Fall", "Spring", "Summer", "Winter")) +
  theme_minimal()

pm25_annual_season

```

Even though this is artifical data - imagine this was actual results. Write a 1 paragraph summary for managers interpreting the patterns you see in the data and plots - and note any implications for public health or policy. You can make it up - as long as it is consistent with the data you generated.

Put your quarto and a rendered version (pdf or html) in your GitHub repository for this assignment. You can use the same repository you created for HW2, but make you name the files clearly (or put in a new directory) so that I can find them. For example, you could put the quarto file in a new directory called ŌĆ£HW3ŌĆØ and name it something like ŌĆ£HW3_air_quality.qmdŌĆØ.

Turn in your *GitHub* repo link on Canvas
