---
title: "homework_3"
format: html
editor: visual
---

## Part 1: Build the Dataset

Create a data frame with 365 rows representing daily air quality readings at a monitoring station. Your data frame must include the following columns:

1.  `day` — a sequence of integers from 1 to 365

2.  `pm25` — daily PM2.5 concentration (µg/m³), sampled from a distribution of your choice using one of the data generation functions covered in class (`runif`, `rnorm`, `rpois`). Choose parameters that produce realistic values — look up typical urban PM2.5 ranges to guide your choices.

3.  `season` — a factor column with levels `"winter"`, `"spring"`, `"summer"`, `"fall"`, assigned based on day number (days 1–90, 91–180, 181–270, 271–365)

4.  `risk` — an empty column initialized to `NA` — you will fill this in Part 2

```{r}

library(dplyr)
library(tidyverse)
#creating data 
day = 1:365
day
pm25 = runif(min=0, max=50, n=365)
pm25

risk = NA


seasons <- case_when(
  day %in% 1:90 ~ "winter",
  day %in% 91:180 ~"spring",
  day %in% 181:270 ~ "summer",
  day %in% 271:365 ~ "fall")
seasons

seasons <- as.factor(seasons)

data <- data.frame(seasons = seasons,
           pm25 = pm25,
           day = day, 
           risk = risk)
data
```

## **Part 2 – Classify Risk with a For Loop**

Using a `for` loop, fill in the `risk` column for each day based on the following PM2.5 thresholds (adapted from EPA AQI guidelines):

-   `"good"` — PM2.5 below 12 µg/m³

-   `"moderate"` — PM2.5 between 12 and 35.4 µg/m³

-   `"unhealthy"` — PM2.5 above 35.4 µg/m³

*Important* while there are many ways to do this - you must use a *for* loop

Within the same loop, also track the longest consecutive streak of `"unhealthy"` days. This will require checking the previous row’s risk classification at each step. Store the final streak length in a variable called `max_unhealthy_strea`

```{r}

min(pm25)
max(pm25)

# setting a named variable for the threshold makes the code easier to read
unhealthy = 35.4
moderate = 12
good = 0
current_streak <- 0
max_unhealthy_streak <- 0

for (i in 1:length(data$pm25)) {
  data$risk[i] <- 
    ifelse(data$pm25[i] > unhealthy, "unhealthy", 
           ifelse(data$pm25[i] >= moderate, "moderate", "good"))
  if (data$risk[i] == "unhealthy") {
    if (i == 1 || data$risk[i - 1] == "unhealthy") {
      current_streak <- current_streak + 1
    } else {
      current_streak <- 1 
      }
  } else {
    current_streak <- 0 
  }
  if (current_streak > max_unhealthy_streak) 
        max_unhealthy_streak <- current_streak 
  }

max_unhealthy_streak

data



```

## **Part 3 – Seasonal Summary with a For Loop**

Write a `for` loop that computes number of days in each season that fall into unhealthy risk category

```{r}
# setting a named variable for the threshold makes the code easier to read
day_unhealthy <- c(winter = 0, spring = 0, summer = 0, fall = 0)


for (i in 1:length(data$day)) {
  if(data$risk[i] == "unhealthy") {
    seasons[i] <- data$seasons[i] 
    day_unhealthy[season_i] <- day_unhealthy[season_i] + 1
  }
}

m <- day_unhealthy 
m

table(data$seasons[data$risk == "unhealthy"])
data
day_unhealthy
day_unhealthy


max_unhealthy_streak

data

```
